---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2025"
subtitle: "Assignment 2 - Due date 01/27/26"
author: "Trudy"
output: pdf_document
geometry: margin=2.54cm
---

## Submission Instructions

You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github.   

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A02_Sp26.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

## R packages

R packages needed for this assignment:"forecast","tseries", and "dplyr". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r}
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=60), tidy=TRUE) 
#Load/install required package here
library(forecast)
library(tseries)
library(dplyr)
library(readxl)
library(openxlsx)
```


## Data set information  

Consider the data provided in the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx" on our **Data** folder. The data comes from the US Energy Information and Administration and corresponds to the December 2025 Monthly Energy Review. The spreadsheet is ready to be used. Refer to the file "M2_ImportingData_XLSX.Rmd" in our Lessons folder for instructions on how to read $.xlsx$ files.

```{r}
#Importing data set
#Importing data set without change the original file using read.xlsx
file_path <- "./Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx"

energy_data1 <- read_excel(
  path = file_path,
  skip = 12,
  sheet = "Monthly Data",
  col_names = FALSE
)

#Now let's extract the column names from row 11
read_col_names <- read_excel(
  path = file_path,
  skip = 10,
  n_max = 1, 
  sheet="Monthly Data",
  col_names=FALSE
)

#Assign the column names to the data set
colnames(energy_data1) <- read_col_names

#Visualize the first rows of the data set
head(energy_data1)

```


## Question 1
You will work only with the following columns: Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption. Create a data frame structure with these three time series only. Use the command head() to verify your data.

```{r}
q1_data <- energy_data1 %>%
  dplyr::select(
    `Total Biomass Energy Production`,
    `Total Renewable Energy Production`,
    `Hydroelectric Power Consumption`
  )

head(q1_data)

```


## Question 2
Transform your data frame in a time series object and specify the starting point and frequency of the time series using the function ts().

```{r}
# Convert data frame to time series object
q1_ts <- ts(
  q1_data,
  start = c(1973, 1),
  frequency = 12
)

# Check the time series object
q1_ts

```


## Question 3
Compute mean and standard deviation for these three series.

```{r}
q3_stats <- data.frame(
  Mean = colMeans(q1_ts),
  Std_Dev = apply(q1_ts, 2, sd)
)

q3_stats
```


## Question 4
Display and interpret the time series plot for each of these variables. Try to make your plot as informative as possible by writing titles, labels, etc. For each plot add a horizontal line at the mean of each series in a different color.

```{r}
library(ggplot2)
library(forecast)

# 1. Total Biomass Energy Production
p1 <- autoplot(q1_ts[, "Total Biomass Energy Production"]) +
  geom_hline(
    yintercept = mean(q1_ts[, "Total Biomass Energy Production"]),
    color = "red",
    linewidth = 1
  ) +
  labs(
    title = "Total Biomass Energy Production Over Time",
    x = "Time",
    y = "Energy Production"
  )
print(p1)

# 2. Total Renewable Energy Production
p2 <- autoplot(q1_ts[, "Total Renewable Energy Production"]) +
  geom_hline(
    yintercept = mean(q1_ts[, "Total Renewable Energy Production"]),
    color = "blue",
    linewidth = 1
  ) +
  labs(
    title = "Total Renewable Energy Production Over Time",
    x = "Time",
    y = "Energy Production"
  )
print(p2)

# 3. Hydroelectric Power Consumption
p3 <- autoplot(q1_ts[, "Hydroelectric Power Consumption"]) +
  geom_hline(
    yintercept = mean(q1_ts[, "Hydroelectric Power Consumption"]),
    color = "darkgreen",
    linewidth = 1
  ) +
  labs(
    title = "Hydroelectric Power Consumption Over Time",
    x = "Time",
    y = "Energy Consumption"
  )
print(p3)
```


## Question 5
Compute the correlation between these three series. Are they significantly correlated? Explain your answer.

```{r}
q5_cor <- cor(q1_ts)

q5_cor

```
The correlation matrix shows a very strong positive correlation between total biomass energy production and total renewable energy production (around 0.97), indicating that these two series move closely together over time. This result is expected since biomass energy constitutes a substantial component of total renewable energy production.

In contrast, hydroelectric power consumption exhibits very weak correlations with both biomass energy production and total renewable energy production, suggesting that hydropower dynamics are largely independent of biomass-related and overall renewable energy trends.

Overall, only biomass energy production and total renewable energy production are strongly and significantly correlated, while hydroelectric power consumption is not meaningfully correlated with the other two series.


## Question 6
Compute the autocorrelation function from lag 1 up to lag 40 for these three variables. What can you say about these plots? Do the three of them have the same behavior?

```{r}
library(forecast)

# ACF: Total Biomass Energy Production
Acf(q1_ts[, "Total Biomass Energy Production"],
    lag.max = 40,
    main = "ACF: Total Biomass Energy Production")

# ACF: Total Renewable Energy Production
Acf(q1_ts[, "Total Renewable Energy Production"],
    lag.max = 40,
    main = "ACF: Total Renewable Energy Production")

# ACF: Hydroelectric Power Consumption
Acf(q1_ts[, "Hydroelectric Power Consumption"],
    lag.max = 40,
    main = "ACF: Hydroelectric Power Consumption")
```
The three time series do not exhibit the same autocorrelation behavior.

Total Biomass Energy Production and Total Renewable Energy Production show very strong and slowly decaying positive autocorrelations across many lags, indicating strong persistence and trend-dominated, non-stationary behavior.

In contrast, Hydroelectric Power Consumption displays an oscillating autocorrelation pattern with alternating positive and negative values and clear peaks at seasonal lags (around 12, 24, and 36), suggesting strong seasonal dynamics rather than a long-term trend.


## Question 7
Compute the partial autocorrelation function from lag 1 to lag 40 for these three variables. How these plots differ from the ones in Q6?

```{r}
library(forecast)

Pacf(q1_ts[, "Total Biomass Energy Production"],
     lag.max = 40,
     main = "PACF: Total Biomass Energy Production (Lag 1–40)")

Pacf(q1_ts[, "Total Renewable Energy Production"],
     lag.max = 40,
     main = "PACF: Total Renewable Energy Production (Lag 1–40)")

Pacf(q1_ts[, "Hydroelectric Power Consumption"],
     lag.max = 40,
     main = "PACF: Hydroelectric Power Consumption (Lag 1–40)")

```
Compared to the ACF plots in Question 6, the PACF plots show a much clearer structure. While the ACF decays slowly over many lags due to strong persistence and trend, the PACF exhibits a dominant spike at lag 1 and significant seasonal spikes around lag 12. This indicates that most of the serial dependence can be explained by short-term (AR(1)) and seasonal effects, rather than long-range direct dependence.

The three series do not exhibit exactly the same behavior. Total Biomass Energy Production and Total Renewable Energy Production show very similar PACF patterns, characterized by strong first-order dependence and seasonal effects. In contrast, Hydroelectric Power Consumption displays stronger seasonal dependence and weaker long-term persistence, reflecting its sensitivity to hydrological cycles rather than long-term growth trends.
